Dependent Object Types

### 2012-09-16 doc/fool/article.tex
exs

Compared to Scala, DOT also simplifies type members. In DOT, a type
member is declared by its lower and upper bounds. Furthermore, unlike
Scala, there is no notion of uninstiated type member. Here is an
example inspired by~\cite{vt}:
\begin{lstlisting}
trait Food
trait Animal {
  type Meal <: Food
  def eat(meal: Meal) {}
}
\end{lstlisting}
In Scala, \code{Meal} is an uninstiated type member of \code{Animal}:
it is not possible to instantiate an \code{Animal} without providing a
concrete binding for \code{Meal}. In DOT, \code{Meal} is a type member
with a lower bound of $\Bot$ and an upper bound of \code{Food}. It is
possible to instantiate an \code{Animal} without further specifying
\code{Meal}, though it would be not possible to feed it. Now,
we define \code{Cow} by refining \code{Animal}.
\begin{lstlisting}
trait Grass extends Food
trait Cow extends Animal {
  type Meal = Grass
}
\end{lstlisting}
In Scala, the type alias \code{type Meal = Grass} defines a concrete
binding for \code{Meal}. Now, we can instantiate \code{Cow}s and feed
them \code{Grass}. In DOT, such a type alias is declared by giving it
identical lower and upper bounds.

@@

\subsection{Scala to DOT}\label{dot-examples-scala-sugar}
// TODO

\subsection{Sugar: Functions}
Like in Scala, we can encode functions as objects with a special
method. We will freely use the following sugar in the remaining of
this paper. Note that the variable $z$ must be fresh.
\begin{align*}
\arrow S T &\iff {\Top \refine z {\mdecl {\mi{apply}} S T}}\\
\fun x S T t &\iff \new z {{\arrow S T}\ldefs{\mi{apply}(x) = t}} z\\
\app f x &\iff \mapp f {\mi{apply}} x\\
\cast T t&\iff \app {(\fun x T T x)} t
\end{align*}

In the remaining examples, for brevity, we will also use $\abs x S t$
for $\fun x S \_ t$ where the return type can be somewhat inferred
from the context.

\subsection{Basics: Booleans, Error, \ldots}
This program defines a $\mi{root}$ object with basic types
($\mi{Unit}$, $\mi{Boolean}$, $\mi{Nat}$) and methods ($\mi{true}$,
$\mi{false}$, $\mi{error}$, $\mi{zero}$). For brevity, we've omitted
the code related to natural numbers. During the object creation, the
method labels such as $\mi{false}$ are all initialized.
\begin{align*}
&\mlnew {\mi{root}} {\Top \mlrefine r {\\\displaybreak[3]
&\ \Ldecl {\mi{Unit}} \Bot \Top\\\displaybreak[3]
&\ \mdecl {\mi{unit}} \Top {r.\mi{Unit}}\\\displaybreak[3]
&\ \Ldecl {\mi{Boolean}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
&\ \gap \ldecl {\mi{ifNat}} \arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {\arrow {(\arrow {r.\mi{Unit}} {r.\mi{Nat}})} {r.\mi{Nat}}}\\\displaybreak[3]
&\gap}}\\\displaybreak[3]
&\ \mdecl {\mi{false}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
&\ \mdecl {\mi{true}} {r.\mi{Unit}} {r.\mi{Boolean}}\\\displaybreak[3]
&\ \mdecl {\mi{error}} {r.\mi{Unit}} \Bot\\\displaybreak[3]
& \ \ldots\\\displaybreak[3]
&}\mlldefs{\\\displaybreak[3]
&\ \mi{unit}(x) = {\new u {\mi{root}.\mi{Unit}} u}\\\displaybreak[3]
%
&\ \mi{false}(u) =\\\displaybreak[3]
&\ \gap \mlnew {\mi{ff}} {\mi{root}.\mi{Boolean} \mlldefs{\\\displaybreak[3]
&\ \gap\gap {\mi{ifNat}} = \abs t {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}} {\\&\gap\gap\gap\abs e {\arrow {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Nat}}}\\&\gap\gap\gap\app e {\mi{root}.\mi{unit}}}\\\displaybreak[3]
&\ \gap}}{
\ \gap \mi{ff}}\\\displaybreak[3]
%
&\ \mi{true}(u) = \ldots\\\displaybreak[3]
%
&\ \mi{error}(u) = {\app {\mi{root}.\mi{error}} u}\\\displaybreak[3]
&\ \ldots\\\displaybreak[3]
&}}{
\ldots
}
\end{align*}

\subsection{Lists}
Polymorphic lists can be expressed using an abstract type member for
the element type ($\mi{Elem}$). We can instantiate a refinement of the
list package to manipulate lists with a particular element type.
\begin{align*}
&\mlnew {\mi{genLists}} {\Top \mlrefine g {\Ldecl {\mi{ListPackage}} \Bot {\Top \mlrefine p {\\\displaybreak[3]
&\ \Ldecl {\mi{Elem}} \Bot \Top\\\displaybreak[3]
&\ \Ldecl {\mi{List}} \Bot {\Top \mlrefine z {\\\displaybreak[3]
&\ \gap \mdecl {\mi{isEmpty}} {\mi{root}.\mi{Unit}} {\mi{root}.\mi{Boolean}}\\\displaybreak[3]
&\ \gap \mdecl {\mi{head}} {\mi{root}.\mi{Unit}} {p.\mi{Elem}}\\\displaybreak[3]
&\ \gap \mdecl {\mi{tail}} {\mi{root}.\mi{Unit}} {p.\mi{List}}\\\displaybreak[3]
&\gap}}\\\displaybreak[3]
&\ \mdecl {\mi{nil}} {\mi{root}.\mi{Unit}} {p.\mi{List}}\\\displaybreak[3]
&\ \mdecl {\mi{cons}} {p.\mi{Elem}} {\arrow {p.\mi{List}} {p.\mi{List}}}\\\displaybreak[3]
&}}}}{
\mlnew {\mi{natLists}} {{\mi{genLists}.\mi{ListPackage}} \mlrefine p {\\\displaybreak[3]
&\gap \Ldecl {\mi{Elem}} {\mi{root}.\mi{Nat}} {\mi{root}.\mi{Nat}} }\mlldefs{\\\displaybreak[3]
&\ {\mi{nil}}(u) = \ldots\\\displaybreak[3]
&\ {\mi{cons}}(hd) = \ldots\\\displaybreak[3]
&}}}{
\ldots
}
\end{align*}

>>>>

Compared to Scala, DOT also simplifies type members. In DOT, a type
member is declared by its lower and upper bounds. Furthermore, unlike
Scala, there is no notion of uninstantiated type member. Here is an
example inspired by~\cite{vt}:
\begin{lstlisting}
trait Food
trait Animal {
  type Meal <: Food
  def eat(meal: Meal) {}
}
\end{lstlisting}
In Scala, \code{Meal} is an uninstantiated type member of \code{Animal}:
it is not possible to instantiate an \code{Animal} without providing a
concrete binding for \code{Meal}. In DOT, \code{Meal} is a type member
with a lower bound of $\Bot$ and an upper bound of \code{Food}. It is
possible to instantiate an \code{Animal} without further specifying
\code{Meal}, though it would be not possible to feed it. Now,
we define \code{Cow} by refining \code{Animal}.
\begin{lstlisting}
trait Grass extends Food
trait Cow extends Animal {
  type Meal = Grass
}
\end{lstlisting}
In Scala, the type alias \code{type Meal = Grass} defines a concrete
binding for \code{Meal}. Now, we can instantiate \code{Cow}s and feed
them \code{Grass}. In DOT, such a type alias is declared by giving it
identical lower and upper bounds.

@@

\subsection{Functions as Sugar}
Like in Scala, we can encode functions as objects with a special
method. Note that the variable $z$ must be fresh.
\begin{align*}
\arrow S T &\iff {\Top \refine z {\mdecl {\mi{apply}} S T}}\\
\fun x S T t &\iff \new z {{\arrow S T}\ldefs{\mi{apply}(x) = t}} z\\
\app f x &\iff \mapp f {\mi{apply}} x\\
\cast T t&\iff \app {(\fun x T T x)} t
\end{align*}

We will freely use the following sugar in the remaining of this
paper. We will also sometimes use $\abs x S t$ for $\fun x S \_ t$,
omitting the return type for convenience and brevity.

\subsection{Class Hierarchies}
A class hierarchy such as
\begin{lstlisting}
object pets {
  trait Pet
  trait Cat extends Pet
  trait Dog extends Pet
  trait Poodle extends Dog
  trait Dalmatian extends Dog
}
\end{lstlisting}
can be easily represented by concrete type members, setting the upper
bounds appropriately:
\begin{align*}
\mlnew {\mi{pets}} {&\Top \mlrefine z {\\
&\ \Ldecl {\mi{Pet_c}} \Bot \Top\\
&\ \Ldecl {\mi{Cat_c}} \Bot {z.{\mi{Pet_c}}}\\
&\ \Ldecl {\mi{Dog_c}} \Bot {z.{\mi{Pet_c}}}\\
&\ \Ldecl {\mi{Poodle_c}} \Bot {z.{\mi{Dog_c}}}\\
&\ \Ldecl {\mi{Dalmatian_c}} \Bot {z.{\mi{Dog_c}}}\\
} \ldefs{}}{}
\end{align*}

The lower of bounds of $\Bot$ ensures that these concrete types are
nominal as the \textsc{$\sub$-tsel} rule cannot be meaningfully
applied.

\subsection{Abstract Type Members}
The \code{choices.Alt} trait takes three abstract type members:
\code{C}, \code{A}, \code{B}. \code{A} and \code{B} are upper bounded
by \code{C}. The intention is that the \code{choose} function takes an
\code{A} and a \code{B} and returns one or the other.
\begin{lstlisting}
object choices {
  trait Alt {
    type C
    type A <: C
    type B <: C
    val choose : A => B => C
  }
}
\end{lstlisting}
In DOT, we can state more precisely the return type of $\mi{choose}$,
thanks to union types:
\begin{align*}
&\new {\mi{choices}} {\Top \mlrefine z {\\
&\gap\Ldecl {\mi{Alt_c}} {\Bot} {\Top \mlrefine a {\\
&\gap\gap\Ldecl C \Bot \Top\\
&\gap\gap\Ldecl A \Bot {a.C}\\
&\gap\gap\Ldecl B \Bot {a.C}\\
&\gap\gap\mdecl {\mi{choose}} {a.A} {\arrow {a.B} {a.A \tor a.B}}\\
&\gap}}\\
&}\ldefs{}}
\end{align*}

Using lower bounds of $\Bot$ for the abstract type members ensures
they can vary covariantly. On the other hand, $\mi{choose}$ cannot be
called meaningfully, unless $A$ and $B$ have realizable lower bounds.

As expected, 
\begin{align*}
&{\mi{choices.Alt_c} \refine a {\Ldecl C \Bot {\mi{pets.Dog_c}}}}\\
\sub\gap&{\mi{choices.Alt_c} \refine a {\Ldecl C \Bot {\mi{pets.Pet_c}}}}
\end{align*}
but
\begin{align*}
&{\mi{choices.Alt_c} \refine a {\Ldecl C {\mi{pets.Dog_c}} {\mi{pets.Dog_c}}}}\\
\not\sub\gap&{\mi{choices.Alt_c} \refine a {\Ldecl C {\mi{pets.Pet_c}} {\mi{pets.Pet_c}}}}
\end{align*}

\subsection{Polymorphic Operators as Sugar}
In Scala, we can implement a polymorphic operator \code{pickLast}
which takes concrete types for \code{C}, \code{A} and \code{B} and
implements a \code{choices.Alt} instance where the \code{choose}
function picks the \code{B} element -- note the precision of the
\code{choose} function which has been refined to return an element of
exactly type \code{B}.
\begin{lstlisting}
def pickLast[Cp,Ap <: Cp,Bp <: Cp] = new Alt {
  type C = Cp
  type A = Ap
  type B = Bp
  val choose: A => B => B = a => b => b
}

val potty = new Poodle {}
val dotty = new Dalmatian {}
val picker = pickLast[Dog,Poodle,Dalmatian]
val p: picker.A = potty
val r: picker.B = picker.choose(potty)(dotty)
\end{lstlisting}

In DOT, we can implement such a polymorphic operator as sugar. Here,
it is not convenient to just return a complex term like we did for
functions as sugar because then invoking $\mi{choose}$ falls under the
$\textsc{term-$\ni$}$ restriction, which doesn't allow the ``self''
type to occur in the result of the method invocation. So the
translation involves explicitly binding an object to the result of the
polymorphic operator. We translate
\begin{align*}
\textbf{val } x^a = \textbf{pickLast}(T^C, T^A, T^B); e^a
\end{align*}
to
\begin{align*}
&\mlnew {x^a} {\mi{choices.Alt_c} \mlrefine {x^a} {\\
&\gap\Ldecl C {T^C} {T^C}\\
&\gap\Ldecl A {T^A} {T^A}\\
&\gap\Ldecl B {T^B} {T^B}\\
&\gap\mdecl {\mi{choose}} {x^a.A} {\arrow {x^a.B} {x^a.B}}\\
&}\ldefs{\mi{choose}(a)=\fun b {x^a.B} {x^a.B} b}}
{e^a}
\end{align*}

Now, given
\begin{align*}
&\mlnew p {\mi{pets.Poodle_c}}
{\mlnew d {\mi{pets.Dalmatian_c}}
{\textbf{val } a = \textbf{pickLast}({\mi{pets.Dog_c}}, {\mi{pets.Poodle_c}}, {\mi{pets.Dalmatian_c}});}}
\end{align*}

The type of $a$ is a subtype of ${\mi{choices.Alt_c}}$:
\begin{align*}
&\cast \Top {\\
&\cast {\mi{choices.Alt_c}} {\\
&\cast {{\mi{choices.Alt_c}} \refine a {\Ldecl C \Bot {\mi{pets.Dog_c}}}} {\\
&a}}}
\end{align*}

The type of $p$ is a subtype of $a.A$:
\begin{align*}
\cast \Top {\cast {a.A} {p}}
\end{align*}

$a$ chooses a $\mi{pets.Dalmatian_c}$:
\begin{align*}
\cast \Top {\cast {\mi{pets.Dalmatian_c}} {\app {a.choose(p)} d}}
\end{align*}

$a$ enforces that it's first argument be a $\mi{pets.Poodle_c}$ and it's second a $\mi{pets.Dalmatian_c}$. The following does not typecheck for this reason:
\begin{align*}
\cast \Top {\app {a.choose(d)} p}
\end{align*}
